Bayesian Nonparametric System Reliability\\ using Sets of Priors

### 2016-01-13 ijar-npb-ms.tex
Section 4 language tweaks -- feel free to roll back any/all of them!

The parametrisation of the Beta distribution used above is common,
and allows to interpret $\alpha^k_t$ and $\beta^k_t$ as
hypothetical numbers of functioning and failed components of type $k$ at time $t$, respectively.
However, for our generalisation to sets of priors,
it is useful to consider a different parametrisation.

For discussing this reparametrisation, we will drop the super- and subscript $k$ and $t$ for a while.
Instead of $\alpha$ and $\beta$, we consider the parameters $\nz$ and $\yz$, where
\begin{align}
\nz &= \alpha + \beta &
&\text{and} &
\yz &= \frac{\alpha}{\alpha + \beta} \,,
\label{eq:reparam}
\end{align}
or vice versa, $\alpha = \nz\yz$ and $\beta = \nz(1-\yz)$.
The upper index ${}\uz$ is used to identify $\nz$ and $\yz$ as prior parameter values,
in contrast to their posterior values $\nn$ and $\yn$
obtained after observing $n$ failure times, see below.
$\nz$ and $\yz$ are sometimes called \emph{canonical} parameters,
identified from rewriting the density in a canonical form;
see, e.g., \cite[pp.~202 and 272f]{2000:bernardosmith}, or \cite[\S 1.2.3.1]{2013:diss-gw}.
This canonical form gives a common structure to all conjugate priors in exponential families.

From the properties of the Beta distribution,
it follows that $\yz = \E[p]$ is the prior expectation for the functioning probability $p$,
and that the higher $\nz$, the more probability weight will be concentrated around $\yz$,
as $\V(p) = \frac{\yz (1-\yz)}{\nz + 1}$.
Furthermore, $\nz$ can be interpreted as a pseudocount or prior strength,
this is clear from the interpretation of $\alpha$ and $\beta$ and \eqref{eq:reparam},
and will be further illustrated below through revisiting the posterior
%based on observed lifetimes $\vec{t} = (t_1, \ldots, t_{n})$.
given that $s$ of $n$ components function.
We have that
$p \mid s$ is Beta distributed with the updated parameters
\begin{align}
\nn &= \nz + n\,, &
\yn &= \frac{\nz}{\nz + n} \cdot \yz + \frac{n}{\nz + n} \cdot \frac{s}{n}\,.
\label{eq:nyupdate}
\end{align}
After seeing that $s$ out of $n$ components function (at time $t$),
the posterior mean $\yn$ for $p$ is a weighted average of
the prior mean $\yz$ and $s/n$ (the fraction of functioning components in the data),
with the weights $\nz$ and $n$, respectively.
We see that $\nz$ plays the same role for the prior mean $\yz$
as the sample size $n$ for the observed mean $s/n$,
thus the notion of pseudocount.
Indeed, the higher $\nz$, the higher the weight for $\yz$
in the weighted average calculation of $\yn$,
so $\nz$ gives the strength of the prior as compared to the sample size $n$.
%The parameters $\nz$ and $\yz$ have a more intuitive interprtation,

The Beta-Binomial probability mass function (pmf) in terms of the updated parameters $\nn$ and $\yn$,
now again with indicators $t$ and $k$,
is therefore
\begin{align}
P(C^k_t = l_k \mid s^k_t) &= {m_k \choose l_k} \frac{B(l_k + \nn_{k,t}\yn_{k,t}, m_k - l_k + \nn_{k,t}(1-\yn_{k,t}))}
                                                    {B(\nn_{k,t}\yn_{k,t}, \nn_{k,t}(1-\yn_{k,t}))} \,,
\label{eq:postpredCny}
\end{align}
and the corresponding cumulative mass function (cmf) is given by
\begin{align}
F_{C^k_t\mid s^k_t}(l_k) &= P(C^k_t \le l_k \mid s^k_t) = \sum_{j_k=0}^{l_k} P(C^k_t = j_k \mid s^k_t)\,.
\label{eq:postpredCnycmf}
\end{align}

The parametrization in terms of prior mean and prior strength
enables us to see that in this conjugate setting,
learning from data amounts to averaging between prior and data.
While this allows for the tractability of the model,
it also comes with a serious drawback:
When observed data are very much different from what is assumed in the prior,
this conflict is simply averaged out,
and is not reflected in the posterior or the posterior predictive.

As a simple example, assume that we expect $\ptk$ to be about $0.75$ for a certain $k$ and $t$,
so we choose $\yktz = 0.75$,
and that we value this choice of mean functioning probability with $\nktz = 8$,
i.e., like having seen $8$ observations with a mean $0.75$.
Assume further that there are $n_k = 16$ components of type $k$ in the test data.
When we observe $s^k_t = 12$ components functioning at time $t$,
so $s^k_t/n_k = 0.75$ as we expect,
then the updated parameters are $\nktn = 24, \yktn=0.75$.
Unexpectedly observing that no component functions at time $t$ instead
leads to parameters $\nktn = 24, \yktn=0.25$.
The prior and the posteriors based on these two observation scenarios
are depicted in Figure~\ref{fig:singleprior-pdc}
through their Beta densities and cdfs (left)
and corresponding Beta-binomial probabilities $P(C^k_t = l_k\mid s^k_t)$ and cmfs
for the case of $m_k = 5$ (right).

Due to symmetry, we see that both posteriors have the same variance or spread,
although arising from two fundamentally different scenarios.
Posterior 1 is based on data exactly according to prior expectations;
the increase in confidence on $\ptk \approx 0.75$
is reflected in a more pointed posterior density,
and the posterior predictive is changed only slightly.
It is quite disturbing to see the same degree of confidence in Posterior 2,
being based on data in sharp conflict with prior expectations.
Posterior 2 places most probability weight around $0.25$,
averaging between prior expectation and data,
with the same variance or spread as Posterior 1.
Instead of acknowledging the conflict between observed functioning probability zero
and expected functioning probability $0.75$ by increased variance,
Posterior 2 gives a false sense of certainty,
also reflected in the posterior predictive,
where both Posterior 1 and 2 have the same variance as well.

\begin{figure}
\includegraphics[width=\textwidth]{singleprior-pdc2}
\caption{Beta densities (top left) and cdfs (bottom left),
with the corresponding Beta-binomial probability mass functions (top right) and cumulative mass functions (bottom right),
for a prior with $\nktz = 8, \yktz = 0.75$,
and posteriors based on $n^k_t=16$ observations with $s^k_t=12$ (Posterior~1) and $s^k_t=0$ (Posterior~2), respectively.
Data for Posterior~1 confirm prior assumptions,
while data for Posterior~2 are in conflict with the prior.
However, this conflict is averaged out,
and Posterior 1 and Posterior 2 have the same spread, both in the posterior pdf/cdf and the posterior predictive pmf/cmf,
such that Posterior 2 gives a false sense of certainty despite the massive conflict between prior and data.}
\label{fig:singleprior-pdc}
\end{figure}

To counter this unwanted behaviour,
we propose to use an imprecise probability approach
based on sets of Beta priors, described next.

>>>>

The parametrisation of the Beta distribution used above is common,
and allows $\alpha^k_t$ and $\beta^k_t$ to be interpreted as
hypothetical numbers of functioning and failed components of type $k$ at time $t$, respectively.
However, when we generalise to sets of priors in the sequel,
it is useful to consider a different parametrisation.

For clarity of presentation we will temporarily drop the super- and subscript $k$ and $t$ indices for component type and time.
Instead of $\alpha$ and $\beta$, we consider the parameters $\nz \in [0, \infty)$ and $\yz \in [0,1]$, where
\begin{align}
\nz &= \alpha + \beta &
&\text{and} &
\yz &= \frac{\alpha}{\alpha + \beta} \,,
\label{eq:reparam}
\end{align}
or equivalently, $\alpha = \nz\yz$ and $\beta = \nz(1-\yz)$.
The upper index ${}\uz$ is used to identify these as prior parameter values,
in contrast to their posterior values $\nn$ and $\yn$
obtained after observing $n$ failure times (see below).
$\nz$ and $\yz$ are sometimes called \emph{canonical} parameters,
identified from rewriting the density in canonical form;
see for example \cite[pp.~202 and 272f]{2000:bernardosmith}, or \cite[\S 1.2.3.1]{2013:diss-gw}.
This canonical form gives a common structure to all conjugacy results in exponential families.

From the properties of the Beta distribution,
it follows that $\yz = \E[p]$ is the prior expectation for the functioning probability $p$,
and that larger $\nz$ values lead to greater concentration of probability measure around $\yz$,
since $\V(p) = \frac{\yz (1-\yz)}{\nz + 1}$.
Consequently, $\nz$ represents the prior strength and moreover can be directly interpreted as a pseudocount, as will become clear.
%this is clear from the interpretation of $\alpha$ and $\beta$ and \eqref{eq:reparam},
%and will be further illustrated below through revisiting the posterior
%%based on observed lifetimes $\vec{t} = (t_1, \ldots, t_{n})$.
Indeed, consider the posterior given that $s$ out of $n$ components function:
by conjugacy
$p \mid s$ is Beta distributed with updated parameters
\begin{align}
\nn &= \nz + n\,, &
\yn &= \frac{\nz}{\nz + n} \cdot \yz + \frac{n}{\nz + n} \cdot \frac{s}{n}\,.
\label{eq:nyupdate}
\end{align}
Thus, after observing that $s$ out of $n$ components function (at time $t$),
the posterior mean $\yn$ for $p$ is a weighted average of
the prior mean $\yz$ and $s/n$ (the fraction of functioning components in the data),
with weights $\nz$ and $n$, respectively.
Therefore $\nz$ takes on the same role for the prior mean $\yz$
as the sample size $n$ does for the observed mean $s/n$,
leading to the notion of it being a pseudocount.
%Indeed, the higher $\nz$, the higher the weight for $\yz$
%in the weighted average calculation of $\yn$,
%so $\nz$ gives the strength of the prior as compared to the sample size $n$.
%The parameters $\nz$ and $\yz$ have a more intuitive interprtation,

Reintroducing time and component type indices, the posterior predictive Beta-Binomial probability mass function (pmf) can be written in terms of the updated parameters as
\begin{align}
P(C^k_t = l_k \mid s^k_t) &= {m_k \choose l_k} \frac{B(l_k + \nn_{k,t}\yn_{k,t}, m_k - l_k + \nn_{k,t}(1-\yn_{k,t}))}
                                                    {B(\nn_{k,t}\yn_{k,t}, \nn_{k,t}(1-\yn_{k,t}))} \,,
\label{eq:postpredCny}
\end{align}
with the corresponding cumulative mass function (cmf) given by
\begin{align}
F_{C^k_t\mid s^k_t}(l_k) &= P(C^k_t \le l_k \mid s^k_t) = \sum_{j_k=0}^{l_k} P(C^k_t = j_k \mid s^k_t)\,.
\label{eq:postpredCnycmf}
\end{align}

The parameterisation in terms of prior mean and prior strength (or pseudocount)
makes clear that in this conjugate setting,
learning from data corresponds to averaging between prior and data.
This form is attractive not only because it enhances the interpretability of the model and prior specification,
but crucially it also makes clear what should be a serious concern in any Bayesian analysis:
when observed data differ greatly from what is expressed in the prior,
this conflict is simply averaged out
and is not reflected in the posterior or posterior predictive distributions.

As a simple example, imagine that we expect $\ptk$ to be about $0.75$ for a certain $k$ and $t$,
so we choose $\yktz = 0.75$,
and that we value this choice of mean functioning probability with $\nktz = 8$,
i.e., equivalently to having seen $8$ observations with a mean $0.75$.
If we observe $n_k = 16$ components of type $k$ in the test data and $s^k_t = 12$ function at time $t$,
then $s^k_t/n_k = 0.75$ as we expect,
so that the updated parameters are $\nktn = 24, \yktn=0.75$.
However, in contrast, unexpectedly observing that no component functions at time $t$ instead
leads to parameters $\nktn = 24, \yktn=0.25$.
The prior and the posteriors based on these two scenarios
are depicted in Figure~\ref{fig:singleprior-pdc}
%through their Beta densities and cdfs (left)
%and corresponding Beta-binomial probabilities $P(C^k_t = l_k\mid s^k_t)$ and cmfs
for the case $m_k = 5$ (right).

Due to symmetry, we see that both posteriors have the same variance,
although arising from two fundamentally different scenarios.
Posterior 1 is based on data exactly according to prior expectations;
the increase in confidence on $\ptk \approx 0.75$
is reflected in a more concentrated posterior density,
and the posterior predictive is changed only slightly.
However, it may be cause for concern to see the same degree of confidence in Posterior 2,
which is based on data that is in sharp conflict with prior expectations.
Posterior 2 places most probability weight around $0.25$,
averaging between prior expectation and data,
with the same variance as Posterior 1.
Accordingly, rather than conveying the conflict between observed
and expected functioning probabilities with increased variance,
Posterior 2 instead gives a false sense of certainty.

\begin{figure}
\includegraphics[width=\textwidth]{singleprior-pdc2}
\caption{Beta densities (top left) and cdfs (bottom left),
with the corresponding Beta-binomial posterior predictive probability mass functions (top right) and cumulative mass functions (bottom right),
for a prior with $\nktz = 8, \yktz = 0.75$,
and posteriors based on $n^k_t=16$ observations with $s^k_t=12$ (Posterior~1) and $s^k_t=0$ (Posterior~2), respectively.
Data for Posterior~1 confirm prior assumptions,
while data for Posterior~2 are in conflict with the prior.
However, this conflict is averaged out,
and Posterior 1 and Posterior 2 have the same spread, both in the posterior pdf/cdf and the posterior predictive pmf/cmf,
such that Posterior 2 gives a false sense of certainty despite the massive conflict between prior and data.}
\label{fig:singleprior-pdc}
\end{figure}

To enable diagnosis of when this undesirable behaviour occurs,
we propose to use an imprecise probability approach
based on sets of Beta priors, described in the following section.
